---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Chloe Alimurong"
date: 2026-09-07
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```
---

## The question

**In this analysis, I look at median home values across counties in Maryland, and
ask which communities we measure most and least reliably. I am interested in seeing the trends over time, especially before and after COVID and counties with more urban areas. Or, if there's a trend amongst demographics such as having a Bachelors degree. I wonder if there's data in specific communities that's least reliable and if I can find potential explanations in those discrepancies.**

---

## Choosing what to measure

```{r}
my_state <- "MD"        
my_variable <- "B25077_001"     # Median home value
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

**I got 6 rows and there are 24 counties in MD, the row count does not match because I just printed the head (default first 6 rows) so I used print(county_data) and checked there are truly 24 rows.**

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

**The county with the largest margin of error relative to its estimate is Somerset county at 9.1% MOE, making it the least reliable county in MD for my variable.**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 1  ~ "High confidence",
    moe_pct < 5 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

**4 counties in MD fall into the 'High' confidence category. 18 fall into the 'Moderate' confidence category. 2 fall into the 'Low' confidence category. I rearranged the threshold to be 1% for high confidence and 5% for moderate since my state is small and there are counties with an even smaller population so they mostly fell into high confidence. **

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```

```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 1  ~ "High confidence",
    moe_pct < 5 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```

** The average population of high-confidence counties is 863,423 people. The low-confidence counties has an average of  21,975 people while the moderate averages at 148,506 people. This pattern tells us that the greater the average population, the greater likelihood of a higher margin of error is for the median home value variable.**

---

## What this means for policy

**As a MD resident, this makes sense that the counties we know the least about/have less accurate information on, are the counties with smaller populations. Somerset county has the highest MOE at 9.1%, I don't really know anyone from there, nor Kent (5.8%) or Garrett (4.7%) county, the top 2 and 3 for highest MOE. These are counties with wild horses, other wildlife, and beaches. The counties we pass through on our way to Ocean City where you can't really see houses. Somerset county, MD is one of the few counties in the Northeastern-ish US that only recently needed fiber (internet cable) built to. Basically, not a lot of people live there. Not enough to have a significant statistics on median home value. This is likely because of sampling variability in results we can't account for in small counties. ACS is a survey of a sample so bigger counties like Baltimore and Montgomery would receive more responses to the survey than a small one like Somerset. There could be a really expensive or really cheap house that could be skewing the data that adds to statistical uncertainty in a small county. For state agencies studying the variable, median home value, smaller counties would be at a disadvantage. For example, if a state agency wants to allocate funding such as decreasing or increasing the construction of affordable housing units or even discounting/increasing APR for homeowner loans, residents of these smaller counties that need funding would be left out since the data likely misrepresented them or left them out completely. For example, Somerset county's estimated median home value is $164,300, for those looking to move to Somerset county, they wouldn't receive much assistance as the predicted value is lower than most counties in MD.**

---

## Optional challenges - For those craving more

::: {.callout-note}
## Delete this box

**None of these are graded.** There are no extra points. A complete, well-written assignment earns a 2 either way.

They're here for anyone who finished the above and found it straightforward. Pick
one (or more) that interests you. 

```{r}

## Propagate the margin of error

county_wide <- county_wide %>%
  mutate(moe_sum = moe_sum(estimate))
         
head(county_wide)

```

1. **Propagate the margin of error.** If you compute a derived value — a percentage,
   a ratio, a sum — its uncertainty depends on the uncertainty of its inputs, and
   you can't just carry the original MOE along. The Census Bureau publishes formulas
   for this, and `tidycensus` implements them: `moe_sum()`, `moe_prop()`,
   `moe_ratio()`. Compute something derived, and get its margin of error right.

2. **Parameterize the report.** Add `params:` to the YAML so the state is an input
   rather than hard-coded, and the same file regenerates for any state. Render it
   from the terminal with `quarto render assignment2.qmd -P state:NY`.

3. **Write a function and iterate.** Wrap the data pull in a function, then use
   `purrr::map_dfr()` to run it across several states at once. Does the
   small-places-are-noisier pattern hold everywhere, or are some states different?

4. **Change the geographic scale.** Run the same variable at `geography = "tract"`
   inside one county. Compare the margins of error to what you saw at county level.
   What does that imply about the resolution at which you can actually trust ACS
   data?

If you try one, add a short section describing what you did and what you found.
:::

---

## Before you submit

::: {.callout-note}
## Delete this box — checklist

- [ ] Your name is in the YAML header
- [ ] Every instruction callout is deleted
- [ ] Every `[bracketed prompt]` is replaced with your own writing
- [ ] No `___` blanks remain anywhere
- [ ] The file is at `assignments/assignment2.qmd`
- [ ] **Session → Restart R**, then render the whole project. It works.
- [ ] Committed and pushed
- [ ] You opened your live site and looked at the published page
:::
